Even-point Multi-loop Unitarity and its Applications: Exponentiation, Anomalies and Evanescence

### 2023-10-12 emu_JHEP_REsubmission/responseToEditor.txt

Dear Editor,

We thank the referee for their feedback yada yada yada

We have addressed all the minor clarifications in addition to the following typos we have identified through an a final read through:

C1] 5.46 and 5.47 swapped the generating functions of H(ij)(kl) and H(ijkl) -- this has been resolved.
C2] 5.23 had a misplaced '+' that has been removed
...

Furthermore, we appreciative the substance and less superficial feedback and have made the following additions to the manuscript and provide the following clarification,

Q1]
Q2]
...

Sincerely,
JJMC, NHP

### 2023-10-12 arXiv_v1/emu.tex
added subdirectory, shared EMU.nb notebook

\documentclass[11pt,preprint]{article}
\pdfoutput=1
\usepackage{jheppub}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{mathrsfs}
\usepackage{amsmath,amssymb}
\usepackage[dvipsnames]{xcolor}
%\usepackage{hyperref}
\usepackage{xspace}
\usepackage{ifdraft}
\usepackage{epstopdf}
\usepackage{tocloft}
\renewcommand{\cftdot}{.}
\renewcommand{\cftsubsecleader}{\cftdotfill{\cftdotsep}}
\usepackage{slashed}
\usepackage{diagbox}
\usepackage{colortbl}
\usepackage{shuffle}
\usepackage{bbold}
\usepackage{enumitem}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{stackengine}
\def\RA{\rlap{\scaleBox{1.6}{$\leftrightarrow$}}}
\def\DA{\smash{\bclap{\scaleBox{1.6}{$\downarrow$}}}}
\def\mystrut{\rule[-2ex]{0ex}{6ex}}

@@

\newcommand{\scaleIntBubbleProbeC}[2]{ {
\begin{tikzpicture}[baseline=-2.5]
\begin{feynman}
\vertex (a1) at (-.6,.3) {#1};
\vertex (a2) at (-.6, -.3) {#2};
\vertex (a4) at (.5, .4) {${}^{\alpha_3=2D/2-3}$};
\vertex (a6) at (.5, -.5) {${}^{\alpha_4=1}$};
\vertex [dot,scale=2](mid1) at (0,0){};
\vertex [dot,scale=1.5,hgrey0](mid2) at (0,0){};
\vertex [dot,scale=2](mid5) at (1,0){};
\vertex [dot,scale=1.5,hgrey0](mid6) at (1,0){};
\vertex (a3) at (1.6, .3) {#2};
\vertex (a4) at (1.6, -.3) {#2};
\diagram{
(a1) --[ultra thick,](mid1),
(a2) --[ultra thick,](mid1),
(mid1) --[ ultra thick,out=30,in=150,min distance=0.3cm](mid5),
(mid1) --[ ultra thick,out=-30,in=-150,min distance=0.3cm](mid5),
(mid5) --[ultra thick,](a3),
(mid5) --[ultra thick,](a4)
};
\end{feynman}
\end{tikzpicture}
}
}

@@

\author{John Joseph M. Carrasco}
\author{and Nicolas H. Pavao}

\affiliation{Department of Physics and Astronomy, Northwestern
  University, Evanston, Illinois 60208, USA}

\title{Even-point Multi-loop Unitarity and its Applications: Exponentiation, Anomalies and Evanescence}
\emailAdd{carrasco@northwestern.edu, pavao@u.northwestern.edu}

\abstract{
We identify novel  structure in newly computed multi-loop amplitudes and quantum actions for even-point effective field theories, including both the nonlinear sigma model (NLSM) and double-copy gauge theories such as Born-Infeld and its supersymmetric generalizations.  
We exploit special properties of all even-point theories towards establishing an efficient unitarity based amplitude construction. 
In doing so, we find evidence that the leading IR divergence of NLSM amplitudes exponentiates when the target space is $\mathbb{CP}^1\cong SU(2)/U(1)$. We then systematically compute the two-loop anomalous behavior of Born-Infeld, and find that the counterterms needed to restore $U(1)$ invariant behavior at loop-level can be constructed via a symmetric-structure double-copy.  
We also demonstrate that the divergent part of the one-minus $(-+++)$ two-loop anomaly vanishes upon introducing an evanescent operator. In addition to these purely photonic counterterms, we verify through explicit calculation that the anomalous matrix elements that violate $U(1)$ duality invariance can be alternatively cancelled by summing over internal $\mathcal{N}=4$ DBIVA superfields. 
Finally we find that $\mathcal{N}=4$ Dirac-Born-Infeld-Volkov-Akulov (DBIVA) amplitudes admit double-copy construction through two-loop order by reproducing our unitarity based result with a double copy between color-dual $\mathcal{N}=4$ super-Yang-Mills  and our two-loop NLSM amplitudes. 
This result supports the possibility of  color-dual representations for NLSM beyond one-loop. 
We conclude with an overview of how $D$-dimensional four-photon counterterms can be constructed in generality with the symmetric-structure double-copy, and outline a convenient way of counting evanescent operators using Hilbert series as generating functions. }

\makeatletter
\gdef\@fpheader{\,}

\begin{document}
\maketitle

@@

The gauge theories that we study in this paper have sufficient generalized gauge freedom to assign contact diagrams to kinematic numerators such that at every multiplicity at tree level one can write the amplitude in terms of $N_g$ that obey the same algebraic relations as the color factors, $C_g$. For color factors that are composed of adjoint structure constants, $f^{abc}$ they must be anti-symmetric around vertices, and be related by Jacobi relations about each edge.   For $SU(N_c)$ we can take $f^{abc} \propto \text{Tr}(T^a[T^b,T^c])$ for arbitrary representation $T$ of the gauge group. We will also consider generalization beyond the adjoint to color-weights that can include $d^{abc}$ and associated algebraic relations.   For $SU(N_c)$ we can take $d^{abc} \propto \text{Tr}[T^a\{T^b,T^c\}]$.  Symmetric-structure color factors obey symmetric algebraic relations rather than anti-symmetric ones.

@@

\subsubsection{Two-loop $\mathcal{N}=4$ DBIVA} \label{sec:DBIvDC2loop}
At two-loop it sufficient to show that replacing the color factors reproduces the integrals we found above with the unitarity construction. Starting with the NLSM integrands in \eqn{eq:NLSM2bub} and \eqn{eq:NLSMostrich}, we make the same replacement of color factors post integration as was done at one-loop.  We carry this out first for the double-bubble, and find that the full $D$-dimensional NLSM integral simplifies dramatically:
\begin{equation}
\begin{aligned}
\scaleIntCsmall \bigg|^{C_g \rightarrow N^{\mathcal{N}=4}_g}  \!\!&= N^{(12|34)}_{\mathcal{N}=4}\left[\frac{(s_{12}I_2^D(k_{12}))^2}{2}\left(\frac{s_{14}-s_{13}}{(D-1)^2}+s_{12}\right)\right]
\\
&+N^{(12|43)}_{\mathcal{N}=4}\left[\frac{(s_{12}I_2^D(k_{12}))^2}{2}\left(\frac{s_{13}-s_{14}}{(D-1)^2}+s_{12}\right)\right]
\\
&=s_{12}^4 (I_2^{D}(k_{12}))^2(t_8 F^4)^{\text{(max)}}_{(1234)}   \equiv \scaleIntCtune{}{hgreen}{}{hgreen} \,.
\end{aligned}
\end{equation}
The cancellation between dimension dependent factors is even more startling for the ostrich-diagram integral:
\begin{equation}
\begin{aligned}
\scaleIntBscalarsmall{1}{2}{3}{4}\bigg|^{C_g \rightarrow N^{\mathcal{N}=4}_g} \!\!&= N^{(12|34)}_{\mathcal{N}=4}\frac{s_{12}}{3} \left[\frac{(D-1)(D-4)s_{14}+2(D-2)^2 s_{13}}{(D-1)(4-3D)}\right][I_3\circ I_2]^D(k_{12})
\\
&+N^{(12|43)}_{\mathcal{N}=4}\frac{s_{12}}{3} \left[\frac{(D-1)(D-4)s_{13}+2(D-2)^2 s_{14}}{(D-1)(4-3D)}\right][I_3\circ I_2]^D(k_{12})
\\
&+N^{([12]|34)}_{\mathcal{N}=4}\left[\frac{s_{12}^2}{3}\frac{D+1}{D-1}\right][I_3\circ I_2]^D(k_{12})
\\
&=\frac{2}{3}s_{12}^3[I_3\circ I_2]^{D}(k_{12})(t_8 F^4)^{\text{(max)}}_{(1234)} \equiv \scaleIntBtune{}{}{}{}{hgreen}{hgreen}{hgreen}{hgreen} \,.
\end{aligned}
\end{equation}
This alone is sufficient to demonstrate the validity of the double-copy, as these integrated quantities are exactly the same as those produced via generalized unitarity of $\mathcal{N}=4$ DBIVA at two-loop in \eqn{eq:N42bub} and \eqn{eq:N4ostrich}. We emphasize that based on our analysis, the double-copy at two loop clearly holds in {any} dimension, as all the $D$-dependent prefactors drop out when replacing the color factors with $\mathcal{N}=4$ numerators. Considering the consistency of these two construction, this also serves as strong evidence for the existence of color-dual representation for two-loop pion integrands. We leave identifying such valid representation as an enticing future direction worth investigation.

@@

\mathcal{M}^{\text{BI+CT}}_{\text{1-loop}}\big|_{\alpha'^6} = \frac{1}{2}\frac{\alpha'^4}{(4\pi)^2}\frac{1}{30}\left[\scaleIntAvectorODD{}{}{}{}{$\mathcal{T}^{4+}$}{$t_8F^4$}+a_{\text{ev.}}\scaleIntAvectorODD{}{}{}{}{$\mathcal{T}^{\text{ev.}}$}{$t_8F^4$}\right]+\text{perms}\,.

@@

For example, consider the two-loop divergence for the $(++++)$ anomalous matrix element in pure BI theory. The 4D helicity structure can be captured by a symmetric structure double-copy between NLSM pion numerators and a local gauge theory contact at $\mathcal{O}(\alpha'^5)$. The $s$-channel numerators for this symmetric-structure double copy are as follows
\begin{equation}\label{eq:symNum2loop}
n_s^{\text{dd,}\pi} = s^2 \qquad n_s^{\text{dd,HD}} = s^5 \frac{[12][34]}{\langle 12\rangle \langle 34\rangle}\,.
\end{equation}
Double-copying these kinematic numerators yields a matrix element of the form:
\begin{equation} \label{eq:2loopCTsymDC}
\mathcal{M}^{\text{BI}+\text{HD}} = \sum_{g\in \Gamma^3} \frac{n_g^{\text{dd,HD}}n^{\text{dd,}\pi}_g}{d_g} = (s^6+t^6+u^6) \frac{[12][34]}{\langle 12\rangle \langle 34\rangle}\,.
\end{equation}
While this construction lacks any algebraic relations between the four-point kinematic-factors, similar symmetric numerators were found at six-point for NLSM, which obey non-trivial algebraic relations \cite{Carrasco:2022jxn}. In addition to the photon counterterms constructed above, symmetric double copy is likewise a natural description of gravitational counterterms. Rather than composing the local vector numerators, $n^{\text{dd,HD}}$, with symmetric NLSM numerators, $n^{\text{dd,}\pi}_g$, they can also be squared -- yielding a gravitational counterterm:
\begin{equation} \label{eq:2loopCTsymDC}
\mathcal{M}^{\text{GR}+\text{HD}} = \sum_{g\in \Gamma^3} \frac{(n^{\text{dd,HD}}_g)^2}{d_g} = (s^9+t^9+u^9)  \frac{[12][34]}{\langle 12\rangle \langle 34\rangle}\,.
\end{equation}
In general, any four-point symmetric vector numerator \cite{Carrasco:2022jxn} can be constructed from two $D$-dimensional gauge theory building blocks,
\begin{equation}
n_s^{\text{dd,vec,1}} = f_{12}f_{34}\,, \qquad n_s^{\text{dd,vec,2}} = f_{1324}\,,
\end{equation}
by composing with the symmetric scalar numerators, $n_s^{\text{dd,1}} = s$ and  $n_s^{\text{dd,2}} = tu$. For example, with these building blocks, symmetric double-copy construction captures the exceptional four graviton amplitude of Ref.~\cite{Boels:2016xhc}, which only considered linear combinations of gauge theory amplitudes rather than gauge theory numerators. We note that a similar construction was recently identified in Ref.~\cite{Bourjaily:2023tcc} beyond four-point, which demonstrated that gravity amplitudes permit double copy construction in terms of gauge-invariant numerators that do not obey any algebraic relations between themselves, much like the symmetric numerators above in \eqn{eq:symNum2loop}. We see further exploring higher multiplicity examples of symmetric double copy as an important direction of future study.

>>>>

\documentclass[11pt,letter]{article}
\usepackage{jheppub}
\usepackage{amsmath,amssymb}
\usepackage[dvipsnames]{xcolor}
%\usepackage{hyperref}
\usepackage{xspace}
\usepackage{ifdraft}
\usepackage{epstopdf}
\usepackage{slashed}
\usepackage{diagbox}
\usepackage{colortbl}
\usepackage{shuffle}
\usepackage{bbold}
\usepackage{enumitem}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{stackengine}
\def\RA{\rlap{\scaleBox{1.6}{$\leftrightarrow$}}}
\def\DA{\smash{\bclap{\scaleBox{1.6}{$\downarrow$}}}}
\def\mystrut{\rule[-2ex]{0ex}{6ex}}

@@

\newcommand{\scaleIntBubbleProbeC}[2]{ {
\begin{tikzpicture}[baseline=-2.5]
\begin{feynman}
\vertex (a1) at (-.6,.3) {#1};
\vertex (a2) at (-.6, -.3) {#2};
\vertex (a4) at (.5, .4) {${}^{\alpha_3=D/2-2}$};
\vertex (a6) at (.5, -.5) {${}^{\alpha_4=1}$};
\vertex [dot,scale=2](mid1) at (0,0){};
\vertex [dot,scale=1.5,hgrey0](mid2) at (0,0){};
\vertex [dot,scale=2](mid5) at (1,0){};
\vertex [dot,scale=1.5,hgrey0](mid6) at (1,0){};
\vertex (a3) at (1.6, .3) {#2};
\vertex (a4) at (1.6, -.3) {#2};
\diagram{
(a1) --[ultra thick,](mid1),
(a2) --[ultra thick,](mid1),
(mid1) --[ ultra thick,out=30,in=150,min distance=0.3cm](mid5),
(mid1) --[ ultra thick,out=-30,in=-150,min distance=0.3cm](mid5),
(mid5) --[ultra thick,](a3),
(mid5) --[ultra thick,](a4)
};
\end{feynman}
\end{tikzpicture}
}
}

@@

\author[1]{John Joseph M. Carrasco,}
\author[1]{Nicolas H. Pavao}

\affiliation[1]{Department of Physics and Astronomy, Northwestern
  University, Evanston, Illinois 60208, USA}

\title{Even-point Multi-loop Unitarity and its Applications: Exponentiation, Anomalies and Evanescence}

\abstract{
We identify novel  structure in newly computed multi-loop amplitudes and quantum actions for even-point effective field theories, including both the nonlinear sigma model (NLSM) and double-copy gauge theories such as Born-Infeld and its supersymmetric generalizations.  
We exploit special properties of all even-point theories towards efficient unitarity based amplitude construction. 
In doing so, we find evidence that the leading IR divergence of NLSM amplitudes exponentiates when the symmetry group is $\mathbb{CP}^1\cong SU(2)/U(1)$. 
We then systematically compute the two-loop anomalous behavior of Born-Infeld, and find that the counterterms needed to restore $U(1)$ invariant behavior at loop-level can be constructed via a symmetric-structure double-copy.  
We also demonstrate that the divergent part of the one-minus $(-+++)$ two-loop anomaly vanishes upon introducing an evanescent operator. In addition to these pure photon counterterms, we verify through explicit calculation that the anomalous matrix elements that violate $U(1)$ duality invariance can be alternatively cancelled by summing over internal $\mathcal{N}=4$ DBIVA superfields. 
Finally we find that $\mathcal{N}=4$ Dirac-Born-Infeld-Volkov-Akulov (DBIVA) amplitudes permit double-copy construction through two-loop order by reproducing our unitarity based result with a double copy between color-dual $\mathcal{N}=4$ super-Yang-Mills  and our two-loop NLSM amplitudes. 
This result supports the possibility of  color-dual representations for NLSM beyond one-loop. 
We conclude with an overview of how $D$-dimensional four-photon counterterms can be constructed in generality with the symmetric-structure double-copy, and outline a convenient way of counting evanescent operators using Hilbert series as generating functions. }

\newpage

\begin{document}
\maketitle
\flushbottom

@@

The gauge theories that we study in this paper have sufficient generalized gauge freedom to assign contact diagrams to kinematic numerators such that at every multiplicity at tree level one can write the amplitude in terms of $N_g$ that obey the same algebraic relations as the color factors, $C_g$. For color factors that are composed of adjoint structure constants, $f^{abc}$ they must be anti-symmetric around vertices, and be related by Jacobi relations about each edge.   For $SU(N_c)$ we can take $f^{abc} \propto \text{Tr}(T^a[T^b,T^c])$ for arbitrary representation $T$ of the gauge group.

We will also consider generalization beyond the adjoint to color-weights that can include $d^{abc}$ and associated algebraic relations.   For $SU(N_c)$ we can take $d^{abc} \propto \text{Tr}[T^a\{T^b,T^c\}]$.  Symmetric-structure color factors obey symmetric algebraic relations rather than anti-symmetric ones.

@@

\subsubsection{Two-loop $\mathcal{N}=4$ DBIVA} \label{sec:DBIvDC2loop}
At two-loop it sufficient to show that replacing the color factors reproduces the integrals we found above with the unitarity construction. Starting with the NLSM integrands in \eqn{eq:NLSM2bub} and \eqn{eq:NLSMostrich}, we make the same replacement of color factors post integration as was done at one-loop.  We carry this out first for the double-bubble, and find that the full $D$-dimensional NLSM integral simplifies dramatically:
\begin{equation}
\begin{aligned}
\scaleIntCsmall \bigg|^{C_g \rightarrow N^{\mathcal{N}=4}_g}  \!\!&= N^{(12|34)}_{\mathcal{N}=4}\left[\frac{(s_{12}I_2^D(k_{12}))^2}{2}\left(\frac{s_{14}-s_{13}}{(D-1)^2}+s_{12}\right)\right]
\\
&+N^{(12|43)}_{\mathcal{N}=4}\left[\frac{(s_{12}I_2^D(k_{12}))^2}{2}\left(\frac{s_{13}-s_{14}}{(D-1)^2}+s_{12}\right)\right]
\\
&=s_{12}^4 (I_2^{D}(k_{12}))^2(t_8 F^4)^{\text{(max)}}_{(1234)}   \equiv \scaleIntCtune{}{hgreen}{}{hgreen} \,.
\end{aligned}
\end{equation}
The cancellation between dimension dependent factors is even more startling for the ostrich-diagram integral:
\begin{equation}
\begin{aligned}
\scaleIntBscalarsmall{1}{2}{3}{4}\bigg|^{C_g \rightarrow N^{\mathcal{N}=4}_g} \!\!&= N^{(12|34)}_{\mathcal{N}=4}\frac{s_{12}}{3} \left[\frac{(D-1)(D-4)s_{14}+2(D-2)^2 s_{13}}{(D-1)(4-3D)}\right][I_3\circ I_2]^D(k_{12})
\\
&+N^{(12|34)}_{\mathcal{N}=4}\frac{s_{12}}{3} \left[\frac{(D-1)(D-4)s_{13}+2(D-2)^2 s_{14}}{(D-1)(4-3D)}\right][I_3\circ I_2]^D(k_{12})
\\
&+N^{([12]|34)}_{\mathcal{N}=4}\left[\frac{s_{12}^2}{3}\frac{D+1}{D-1}\right][I_3\circ I_2]^D(k_{12})
\\
&=\frac{2}{3}s_{12}^3[I_3\circ I_2]^{D}(k_{12})(t_8 F^4)^{\text{(max)}}_{(1234)} \equiv \scaleIntBtune{}{}{}{}{hgreen}{hgreen}{hgreen}{hgreen} \,.
\end{aligned}
\end{equation}
This alone is sufficient to demonstrate the validity of the double-copy, as these integrated quantities are exactly the same as those produced via generalized unitarity of $\mathcal{N}=4$ DBIVA at two-loop in \eqn{eq:N42bub} and \eqn{eq:N4ostrich}. We emphasize that based on our analysis, the double-copy at two loop clearly holds in {any} dimension, as all the $D$-dependent prefactors drop out when replacing the color factors with $\mathcal{N}=4$ numerators. Considering the consistency of these two construction, this also serves as strong evidence for the existence of color-dual representation for two-loop pion integrands. We leave identifying such valid representation as an enticing future direction worth investigation.

@@

\mathcal{M}^{\text{BI+CT}}_{\text{1-loop}}\big|_{\alpha'^6} = \frac{1}{2}+ \frac{\alpha'^4}{(4\pi)^2}\frac{1}{30}\left[\scaleIntAvectorODD{}{}{}{}{$\mathcal{T}^{4+}$}{$t_8F^4$}+a_{\text{ev.}}\scaleIntAvectorODD{}{}{}{}{$\mathcal{T}^{\text{ev.}}$}{$t_8F^4$}\right]+\text{perms}\,.

@@

For example, consider the two-loop divergence for the $(++++)$ anomalous matrix element in pure BI theory. The 4D helicity structure can be captured by a symmetric structure double-copy between NLSM pion numerators and a local gauge theory contact at $\mathcal{O}(\alpha'^5)$. The $s$-channel numerators for this symmetric-structure double copy are as follows
\begin{equation}\label{eq:symNum2loop}
n_s^{\text{dd,}\pi} = s^2 \qquad n_s^{\text{dd,HD}} = s^5 \frac{[12][34]}{\langle 12\rangle \langle 34\rangle}\,.
\end{equation}
Double-copying these kinematic numerators yields a matrix element of the form:
\begin{equation} \label{eq:2loopCTsymDC}
\mathcal{M}^{\text{BI}+\text{HD}} = \sum_{g\in \Gamma^3} \frac{n_g^{\text{dd,HD}}n^{\text{dd,}\pi}_g}{d_g} = (s^6+t^6+u^6) \frac{[12][34]}{\langle 12\rangle \langle 34\rangle}\,.
\end{equation}
While this construction lacks any algebraic relations between the four-point kinematic-factors, similar symmetric numerators were found at six-point for NLSM, which obey non-trivial algebraic relations \cite{Carrasco:2022jxn}. In addition to the photon counterterms constructed above, symmetric double copy is likewise a natural description of gravitational counterterms. Rather than composing the local vector numerators, $n^{\text{dd,HD}}$, with symmetric NLSM numerators, $n^{\text{dd,}\pi}_g$, they can also be squared -- yielding a gravitational counterterm:
\begin{equation} \label{eq:2loopCTsymDC}
\mathcal{M}^{\text{GR}+\text{HD}} = \sum_{g\in \Gamma^3} \frac{(n^{\text{dd,HD}}_g)^2}{d_g} = (s^9+t^9+u^9)  \frac{[12][34]}{\langle 12\rangle \langle 34\rangle}\,.
\end{equation}
In general, any four-point symmetric vector numerator can constructed from two $D$-dimensional gauge theory building blocks \cite{Carrasco:2022jxn},
\begin{equation}
n_s^{\text{dd,vec,1}} = f_{12}f_{34}\,, \qquad n_s^{\text{dd,vec,2}} = f_{1324}\,,
\end{equation}
by composing with the symmetric scalar numerators, $n_s^{\text{dd,1}} = s$ and  $n_s^{\text{dd,2}} = tu$. With these building blocks, symmetric double copy construction captures the exceptional four graviton amplitude of Ref.~\cite{Boels:2016xhc}, which only considered linear combinations of gauge theory amplitudes rather than gauge theory numerators. We note that a similar construction was recently identified in Ref.~\cite{Bourjaily:2023tcc} beyond four-point, which demonstrated that gravity amplitudes permit double copy construction in terms of gauge-invariant numerators that do not obey any algebraic relations between themselves, much like the symmetric numerators above in \eqn{eq:symNum2loop}. We see further exploring higher multiplicity examples of symmetric double copy as an important direction of future study.
